Phyloseq Preprocessing

### 2021-11-10 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Update microbiome workflow.

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
    if(is.numeric(x)) as.factor(x) else x
})
str(df3[,1:15])
# Visually check that df2 and df3 appear the same; pass check

@@

```{r Save ln biocrates data}
write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.csv")
saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
```

@@

```{r Make Phyloseq object- log biocrates BD1}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

@@

# Save the phyloseq object
save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211105.RData")
```

@@

# Load phenotype data HERE 
df <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
colnames(df[,1:10])

# Trim df to subjects WITH TMAO (primary variable)
df <- df[which(complete.cases(df$tmao) == TRUE),]

# Take a subset
df <- subset(df, select = c( pa_ee_total, endo_ln_rhi, endo_al75, betaine, carnitine, tmao, choline, creatinine, phosphocholine, hei_asa24_totalveg, hei_asa24_totalproteinfoods, hei_asa24_sodium, hei_asa24_refinedgrains, hei_asa24_sfat, hei_asa24_addsug, hei_asa24_wholegrain, hei_asa24_totalscore ))

@@

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
    if(is.numeric(x)) as.factor(x) else x
})
str(df3[,1:15])
# Visually check that df2 and df3 appear the same; pass check

@@

@@

```{r Save ln biocrates data}
write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.csv")
saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
```

@@

```{r Make Phyloseq object- specific TMAO variables}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

@@

# Save the phyloseq object
save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211105.RData")
```

>>>>

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign below if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a data frame and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])

# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
  if(is.character(x)) as.factor(x) else x
})
# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
str(df3[,1:15])

# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
df3.5 <- df3
df3.5[] <- lapply(df3, function(x) {
  relevel(x, ref = "below")
})
str(df3.5)

# Merge categorical data with numeric data
df4 <- merge(df, df3.5, by = 0)
rownames(df4) <- df4[,1]
df4 <- df4[,-1]

@@

```{r Save ln biocrates data}
write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.csv")
saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
```

@@

```{r Make Phyloseq object- log biocrates BD1}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

@@

# Save the phyloseq object
save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211108.RData")
```

@@

# Load phenotype data HERE 
df_tmao <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
colnames(df_tmao[,1:10])

# Trim df to subjects WITH TMAO (primary variable)
df_tmao <- df_tmao[which(complete.cases(df_tmao$tmao) == TRUE),]

# Take a subset
df <- df_tmao
df <- subset(df, select = c( pa_ee_total, endo_ln_rhi, endo_al75, betaine, carnitine, tmao, choline, creatinine, phosphocholine, hei_asa24_totalveg, hei_asa24_totalproteinfoods, hei_asa24_sodium, hei_asa24_refinedgrains, hei_asa24_sfat, hei_asa24_addsug, hei_asa24_wholegrain, hei_asa24_totalscore ))

@@

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
  if(is.character(x)) as.factor(x) else x
})
# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
str(df3[,1:15])

# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
df3.5 <- df3
df3.5[] <- lapply(df3, function(x) {
  relevel(x, ref = "below")
})
str(df3.5)

@@

# Assign 'Description' (subject ID in XXXX terms) to row names
rownames(md2) <- md2$Description

# Make pathologic variables based on TMAO
md2$TMAO_5 <- ifelse(md2$tmao <= 5, "below5", "above5")
md2$TMAO_5 <- as.factor(md2$TMAO_5)
md2$TMAO_5 <- relevel(md2$TMAO_5, ref = "below5")
# 6
md2$TMAO_6 <- ifelse(md2$tmao <= 6, "below6", "above6")
md2$TMAO_6 <- as.factor(md2$TMAO_6)
md2$TMAO_6 <- relevel(md2$TMAO_6, ref = "below6")
# 7
md2$TMAO_7 <- ifelse(md2$tmao <= 7, "below7", "above7")
md2$TMAO_7 <- as.factor(md2$TMAO_7)
md2$TMAO_7 <- relevel(md2$TMAO_7, ref = "below7")

# Lastly, add age, sex, bmi_final to dataframe
df_bins <- subset(df_tmao, select = c(sex, age, bmi_final))
# Merge
md3 <- merge(md2, df_bins, by=0)
row.names(md3) <- md3[,"Row.names"]
md3 <- subset(md3, select = -c(Row.names))
```

@@

```{r Save ln biocrates data}
write.csv(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.csv")
saveRDS(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
```

@@

```{r Make Phyloseq object- specific TMAO variables}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

@@

# Save the phyloseq object
save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData")
```
